Deep Learning and Bayesian Modeling

### 2021-05-23 DeepLearningAndBayesianSoftware.Rmd
getting close to release

---
title: "Deep Learning and Bayesian Modeling: Comparing research use via citation counting"
author: "Breck Baldwin"
date: "5/19/2021"
output:
 bookdown::tufte_html2: default
bibliography: citations.bib
link-citations: yes
---

@@

Stan, a Bayesian modeling language, was released in 2012 to considerable fanfare. A shiny new inference algorithm, HMC with NUTS (@hoffman2014no), promised and delivered fits that could not be fit before. And then deep learning happened. The relese of TensorFlow (@abadi2016tensorflow) opened the doors of deep learning to everyone and in 2016 began garnering significant traction. One result of this was people often assumed that deep learning's successes usurped Bayesian modeling's domain. This was not in our imagination, NSF reviews came back dismissing Stan funding because all the interesting work was assumed to be happening with deep learning. Mind you, deep learning and Bayesian modeling are conceptual cousins but very different from each other and are better thought of as complementary than as antagonistic. Yet Bayesians found themselves in deep learning's shadow somehow. It is 2021 and this document does some very simple analysis around use of Bayesian and deep learning packages as evidenced in the research literature and downloads to get a perspective on what actually happened and is happening. The comparison aims to approach the following goals with very simple research citation metrics:

@@

So what is going on here? Keeping speculation to a minimum, the next question is what does the the distribution look like for non-computer science subject areas?  Below is a table of subject counts for packages starting in 2006 when PyMC was first released to 2021 with relative percentages listed:

@@

allDataDf <- cbind(displayBayesVdeepLearnDf,displayDetailDf)

kable(allDataDf, booktabs = TRUE) %>% kable_styling(font_size = 10)

@@

Table 1 shows the basic subject split between Bayesian modeling and deep learning with total for the two categories, following are the individual package counts with a total as well. Note that the totals do not match because an article can match more than one package and as result gets contributes to the count of multiple packages. On average an article's journal is classified into two subject categories.

Staring with the subject that got us here we see that computer science has 728 Bayesian publications vs 30,029 deep learning publications for 98% of the split. The left most two columns are the same queries used for the graph in Fig 1 with the total in column 3. Columns 4-8 are indivudual packages with percentage of the total in the 9th column.  One sees that TensorFlow has 18,178 references with 51% of the count--note that the total counts are higher because an article can matched by more that one query, each package gets a query, so the article count is 3 if Keras, PyMC and TensorFlow queries match the references and it has a computer science subject classification.

Looking at the tea leaves I leave to others but I'll make the observation that Bayesians are playing a pretty big game in a lot of areas of science and as someone interested in getting funding to Bayesian software this is a pretty strong argument that we are highly relevant.

@@

Below are the queries used to identify the packages. I would estimate a 5% false positive rate based based on informal examination of return sets. For example the stand alone query 'rstan' had the addition of rejecting articles that mentioned 'mit' because there was a common reference to 'http://wwwmath.mit.edu/~rstan/ec/' that existed prior to 2012 when Stan was released. The name 'Stan' presented obvious difficulties and the combination shown achieved desired count drop off at 2012 with reasonable precision and each of the 'OR' terms contributed to coverage significantly. How the word 'mc-stan.org' was tokenized is unknown but there were no results before 2013 for the bare query.

@@

PyMC presented some challenges in that the string PyMC was referenced in early publications but also is ambiguous with chemical terms, so like Stan, author names were included to restrict search as was the case with 'brms'. These queries were also developed searching entire documents rather than the current restriction to the references section of papers. I would expect the false positive rate to be even lower.

## Technical details and access to source/scripts

Supporting technology for this document includes the R language, the tidyverse group of packages that give data and visualization tools and finally this Rmarkdown document developed using the Rstudio IDE (Integrated Development Environment). I intend this to be an evolving document with periodic updates at it's github repo at [https://github.com/breckbaldwin/bayesianImpactMetrics](https://github.com/breckbaldwin/bayesianImpactMetrics) where it lives as `index.Rmd`, is rendered as `index.html` and is viewable as [https://breckbaldwin.github.io/bayesianImpactMetrics/](https://breckbaldwin.github.io/bayesianImpactMetrics/). All source code is viewable in the .Rmd document.

>>>>

---
title: "Deep Learning and Bayesian Modeling"
subtitle: "Comparing research use via citation counting"
author: "Breck Baldwin"
email: "breckbaldwin@gmail.com"
date: "5/23/2021"
output:
 bookdown::html_document2:
   includes:
     in_header: _html/ga.html 
bibliography: citations.bib
link-citations: yes
---

@@

# Introduction

Stan, a Bayesian modeling language, was released in 2012 to considerable fanfare. A shiny new inference algorithm, HMC with NUTS (@hoffman2014no), promised and delivered fits that could not be fit before. And then deep learning happened. The relese of TensorFlow (@abadi2016tensorflow) opened the doors of deep learning to everyone and in 2016 began garnering significant traction. One result of this was people often assumed that deep learning's successes usurped Bayesian modeling's domain. This was not in our imagination, NSF reviews came back dismissing Stan funding because all the interesting work was assumed to be happening with deep learning. Mind you, deep learning and Bayesian modeling are conceptual cousins but very different from each other and are better thought of as complementary than as antagonistic. Yet Bayesians found themselves in deep learning's shadow somehow. It is 2021 and this document does some very simple analysis around use of Bayesian and deep learning packages as evidenced in the research literature to get a perspective on what actually happened and is happening. The comparison aims to approach the following goals with very simple research citation metrics:

@@

So what is going on here? The next question is what does the the distribution look like for non-computer science subject areas?  Below is a table of subject counts for packages starting in 2006 when PyMC was first released to 2021 with relative percentages listed:

@@

allDataDf <- cbind(displayBayesVdeepLearnDf,displayDetailDf)
allDataDf <- allDataDf[,c(1,2,3,11,6,7,10,9,4,8,5,12,13)]
displayedColNames <- colnames(allDataDf)
colnames(allDataDf) <- c(displayedColNames[1:12],'detail totals')
kable(allDataDf, booktabs = TRUE) %>% kable_styling(font_size = 10)

@@

Table 1 shows the basic subject split between Bayesian modeling and deep learning with total for the two categories, following are the individual package counts with a 'detail totals' as well for the individual packages. Note that the totals do not match because an article can match more than one package and as result gets contributes to the count of multiple packages. On average an article's journal is classified into two subject categories.

Staring with the subject that got us here we see that computer science has 728 Bayesian publications vs 29,301 deep learning publications for 98% of the share of 30,029 total. The left most two columns are the same queries used for the graph in Fig 1 with the total in column 3. Columns 4-12 are individual packages with percentage of the total in the 13th column.  Sticking with computer science, one sees that TensorFlow has 18,178 references with 51% of the count--note that the total counts are higher because an article can matched by more that one query, each package gets a query, so the article count is 3 if Keras, PyMC and TensorFlow queries match the references and it has a computer science subject classification.

I leave the tea leaf interpretation to others but I'll make the observation that Bayesians are playing a pretty big game in a lot of areas of science and as someone interested in getting funding to Bayesian software this is a pretty strong argument that we are highly relevant.

@@

Below are the queries used to identify the packages. Direct citation linking does not work well with these packages since many of them had no journal publication to cite in a referring article. As a result a string search was used to identify the citations in the reference section as determined by Scopus.com's indexing engine. I would estimate a 5% false positive rate based based on informal examination of return sets.

For an example of how the queries were developed, the stand alone query 'rstan' had the addition of rejecting articles that mentioned 'mit' because there was a common reference to 'http://wwwmath.mit.edu/~rstan/ec/' that existed prior to 2012 when Stan was released. The name 'Stan' presented obvious difficulties and the combination shown achieved desired count drop off at 2012 with reasonable precision and each of the 'OR' terms contributed to coverage significantly. How the word 'mc-stan.org' was tokenized is unknown but there were no results before 2013 for the solo query.

@@

PyMC presented some challenges in that the string PyMC was referenced in early publications but also is ambiguous with chemical terms, so like Stan, author names were included to restrict search as was the case with 'brms'. These queries were initially developed to search entire documents rather than the current restriction to the references section of papers. I would expect the false positive rate to be even lower.

# Access to source/scripts and acknowledgements

Supporting technology for this document includes the R language (@Rcite), the 'tidyverse' group of packages (@tidyverse) that give data and visualization tools and this Rmarkdown document (@rmarkdown) rendered with 'bookdown' (@bookdown) developed using the Rstudio (@rstudio) IDE (Integrated Development Environment). Supporting package include 'kable-extra' (@kable-extra), a table formatter, the 'scales' (@scales) package for converting to percentages, 'ggrepel' (@ggrepel) for labeling line graphs with dynamically shifting labels, 'jsonlite' (@jsonlite) for JSON data parsing, 'httr' (@httr) for GET requests, 'stringr' (@stringr) for regular expression matching and 'redux' (@redux) for access to the Redis server (@redis) which I highly recomend using as a caching layer for this sort of project.

I'd like to thank the Andrew Gelman's Laplace lab at Columbia University for support.

I intend this to be an evolving document with periodic updates at it's github repo at [https://github.com/breckbaldwin/bayesianImpactMetrics](https://github.com/breckbaldwin/bayesianImpactMetrics) where it lives as `DeepLearningAndBayesianSoftware.Rmd`, is rendered as `DeepLearningAndBayesianSoftware.html` and is viewable as [https://breckbaldwin.github.io/bayesianImpactMetrics/DeepLearningAndBayesianSoftware.html](https://breckbaldwin.github.io/bayesianImpactMetrics/DeepLearningAndBayesianSoftware.html). All source code is viewable in the .Rmd document.

I can be reached at breckbaldwin@gmail.com or via the github issues on the repository.

# References
